---
title: "Global Negation Strategies: A Comparative Analysis"
description: |
  An  investigation into the distribution of negation strategies using Grambank and WALS datasets.
author:
  - name: Holly Lloyd
date : 2026-03-27
output:
  distill::distill_article:
    toc: true
    toc_depth: 2
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)

#install.packages("scales")
library(knitr)
library(scales)
library(patchwork)
library(sf)
library(tidyverse)
library(ggplot2)
library(broom)
# Loading in packages

Gb_values <- read.csv("values.csv")
Gb_languages <- read.csv("languages.csv")
Gb_parameters <- read.csv("parameters.csv")
wals_values <- read.csv("WALSvalues.csv")
wals_languages <- read.csv("WALSlanguages.csv")
# Loading data 
```
# Section 1: Grambank Areal and Genealogical Distribution

The following analysis examines whether negation strategies are primarily driven by geographic proximity (areal diffusion) or family lineage (genealogical inheritance). I focus on two key features GB107 (affixal/clitic negation) and GB137 (clause-final negation).

# 1.1 Chi-Squared Tests for Areal Effect

In order to evaluate the areal diffusion, I utilised Pearson's Chi-squared tests. These tests enable me to determine whether the presence of a negation strategy is significantly associated with specific geographic macroareas. 

```{r}
negation_params <- Gb_parameters[Gb_parameters $ ID %in% c("GB107", "GB137"),]
Gb_negation_values <- Gb_values[Gb_values $ Parameter_ID %in% c("GB107", "GB137"),]
# Filtering for specific parameters

Gb_negations <- merge(Gb_negation_values, Gb_languages,
                      by.x = "Language_ID", by.y = "ID")
# Integrating values with metadata

Gb_wide <- reshape(Gb_negations[, c("Language_ID", "Parameter_ID", "Value",
                                    "Family_name", "Macroarea", "Longitude", "Latitude")],
                   idvar = c("Language_ID", "Longitude", "Latitude"),
                   timevar = "Parameter_ID",
                   direction = "wide")
# Reshaping to wide and specifying column names

areal_table <- table(Gb_wide $ Value.GB137, Gb_wide  $ Macroarea.GB137)
areal_table_chi <- chisq.test(areal_table)
tidy(areal_table_chi) %>%
  mutate(p.value = scales::scientific(p.value)) %>%
  knitr::kable(caption = "Areal Association: GB137",
               col.names = c("Statistic", "P-Value", "DF", "Method"))

areal_table2 <- table(Gb_wide $ Value.GB107, Gb_wide  $ Macroarea.GB107)
areal_table_chi2 <- chisq.test(areal_table2)
tidy(areal_table_chi2) %>%
  mutate(p.value = scales::scientific(p.value)) %>%
  knitr::kable(caption = "Areal Association: GB107",
               col.names = c("Statistic", "P-Value", "DF", "Method"))
# Chi-squared to evaluation associations
```
# 1.2 Genealogical Stability

While areal factors are significant, many language families exhibit vertical stability. I defined this demonstration of strong inheritance as a language family with 100% internal consistency for the analysed features. 

```{r}
family_table <- table(Gb_wide $ Value.GB137, Gb_wide $ Family_name.GB137)
family_props <- prop.table(family_table, margin = 2)

strong_inheritance <- family_props["1", family_props["1", ] == 1.0]
as.data.frame(head(strong_inheritance)) %>%
  knitr::kable(caption = "Strong Inheritance Families: GB137")

family_table2 <- table(Gb_wide $ Value.GB107, Gb_wide $ Family_name.GB107)
family_props2 <- prop.table(family_table2, margin = 2)

strong_inheritance2 <- family_props2["1", family_props2["1", ] == 1.0]
as.data.frame(head(strong_inheritance2)) %>%
  knitr::kable(caption = "Strong Inheritance Families: GB107")
# Investigating genealogical stability

feature_comp <- table(Gb_wide $ Value.GB107, Gb_wide $ Value.GB137)
feature_test <- chisq.test(feature_comp, simulate.p.value = TRUE)
tidy(feature_test) %>%
  knitr::kable(caption = "Correlation Test: GB137 vs. GB107" )
# Testing for correlation between features
```

```{r}
plotGB137 <- na.omit(Gb_wide[, c("Macroarea.GB137", "Value.GB137")])
plotGB137 <- plotGB137[plotGB137 $ Value.GB137 %in% c("0", "1"),]
#Preparing data for GB137 plot

p1 <- ggplot(plotGB137, aes(x = Macroarea.GB137, fill = as.factor(Value.GB137))) +
  geom_bar(position = "fill", width = 0.7) +
  scale_y_continuous(labels = percent) +
  scale_fill_manual(values = c("#56B4E9", "#0072B2"),
                    labels = c("Absent", "Present")) +
  labs(title = "Clause-Final Negation (GB137)",
       subtitle = "Syntactic Position: High regional variation suggests areal diffusion",
       y = "Percent of Languages", x = NULL, fill = "Presence") +
  theme_minimal(base_size = 11) +
  theme(legend.title = element_blank()) +
  coord_flip()
# Creating GB137 plot

plotGB107 <- na.omit(Gb_wide[, c("Macroarea.GB107", "Value.GB107")])
plotGB107 <- plotGB107[plotGB107 $ Value.GB107 %in% c("0", "1"),]
# Preparing data for GB107 plot

p2 <- ggplot(plotGB107, aes(x = Macroarea.GB107, fill = as.factor(Value.GB107))) +
  geom_bar(position = "fill", width = 0.7) +
  scale_y_continuous(labels = percent) +
  scale_fill_manual(values = c("#E69F00", "#D55E00"),
                    labels = c("Absent", "Present")) +
  labs(title = "Affixal/Clitic Negation (GB107)",
       subtitle = "Morphological Strategy: Broader distribution suggests genealogical stability",
       y = "Percent of Languages", x = NULL, fill = "Presence") +
  theme_minimal(base_size = 11) +
  theme(legend.title = element_blank()) +
  coord_flip()
# Creating GB107 plot

combined_plot <- p1 / p2 +
  plot_annotation(title = "Comparative Analysis of Negation Strategies",
                  caption = "Data: Grambank 2026") +
  plot_layout(guides = "collect")
combined_plot
# Combining plots using patchwork
```
# Section 2: WALS Structural Correlation

In contrast to the binary granularity of Grambank, the WALS dataset allows for the investigation into hte dependency between clausal symmetry (113A) and syntactic Position (143G).

```{r}
wals_full <- merge(wals_values, wals_languages,
                   by.x = "Language_ID", by.y = "ID")
# Integrating WALS values with metadata

wals_113a <- wals_full[wals_full $ Parameter_ID == "113A", ]
wals_143g <- wals_full[wals_full $ Parameter_ID == "143G", ]
# Filtering for relevant parameters

wals_areal_table <- table(wals_113a $ Value, wals_113a $ Macroarea)
wals_chi <- chisq.test(wals_areal_table, simulate.p.value = TRUE)
tidy(wals_chi) %>%
  knitr::kable(caption = "Areal Association: 113A vs. Macroarea", 
col.names = c("Statistic", "P-Value", "DF", "Method"))

wals_areal_table2 <- table(wals_143g $ Value, wals_143g $ Macroarea)
wals_chi2 <- chisq.test(wals_areal_table2, simulate.p.value = TRUE)
tidy(wals_chi2) %>%
  knitr::kable(caption = "Areal Association: 1143G vs. Macroarea", 
col.names = c("Statistic", "P-Value", "DF", "Method"))
# Chi-squared for WALS features

wals_subset <- wals_full[wals_full $ Parameter_ID %in% c("113A", "143G"), ]
wals_wide <- reshape(wals_subset[, c("Language_ID", "Parameter_ID", "Value",
                                     "Family", "Macroarea")],
                     idvar = "Language_ID",
                     timevar = "Parameter_ID",
                     direction = "wide")
# Reshaping subset for correlation testing

wals_wide_clean <- na.omit(wals_wide[, c("Value.113A", "Value.143G")])
wals_wide_clean <- droplevels(wals_wide_clean)
# Cleaning data 
```
# 2.1 Strategy vs. Word Order

The contingency table below maps the frequency of specific strategy-order pairings. A significant chi-squared result suggests that certain syntactic positions favor more complex grammatical pathways.

```{r}
combined_wals_table <- table(wals_wide_clean $Value.113A, wals_wide_clean $ Value.143G)
# Creating contingency table

wals_corr <- chisq.test(combined_wals_table, simulate.p.value = TRUE)
# Performing Chi-squared
```

```{r}
knitr::kable(combined_wals_table,
             caption = "Contingency Table: WALS 113A vs WALS 143G")
# Displaying table

tidy(wals_corr) %>%
  knitr::kable(
    caption = "Chi-squared Test Results: Correlation between Strategy and Order",
    col.names = c("Statistic", "P-Value", "DF", "Method"))
# Displaying Chi-squared
```

```{r}
wals_counts <- as.data.frame(table(wals_wide_clean $ Value.113A,
                                   wals_wide_clean $ Value.143G))
colnames(wals_counts) <- c("Strategy", "Order", "Frequency")
# Preparing data for bar chart


wals_counts $ Strategy <- factor(wals_counts $ Strategy,
                                 labels = c("Symmetric", "Asymmetric", "Asymmetric (Auxiliary)"))
wals_counts $ Order <- factor(wals_counts $ Order,
                              labels = c("V-Neg", "Neg-V"))
# Relabeling 

ggplot(wals_counts, aes(x = Order, y = Frequency, fill = Strategy)) +
  geom_bar(stat = "identity", position = "fill") +
  scale_y_continuous(labels = percent) +
  scale_fill_manual(values = c("#F0E442", "#E69F00", "#D55E00")) +
  labs(title = "Distribution of Negation Strategies by Word Order",
       subtitle = "Proportional analysis of WALS features 113A and 143G",
       caption = "Data: WALS 2026", 
       x = "Word Order (143G)", y = "Percentage of Languages", fill = "Strategy (113A)") +
  theme_minimal()
# Creating bar graph
```

```{r}
world_map <- map_data("world")
# retrieving map coordinates

WALS_map <- ggplot() +
  geom_polygon(data = world_map, aes(x = long, y = lat, group = group),
               fill = "grey95", colour = "grey80", linewidth = 0.1) +
  geom_point(data = wals_113a, aes(x = Longitude, y = Latitude,
                                   colour = as.factor(Value)),
             size = 0.8, alpha = 0.7) +
  scale_colour_manual(values = c("#D55E00", "#E69F00", "#0072B2", "#56B4E9")) +
  coord_fixed(1.3) +
  theme_void() +
  theme(legend.position = "none") +
  labs(title = "WALS 113A: Symmetric and Asymmetric Negation")
# Creating WALS 113A map

Gb_map <- ggplot() +
  geom_polygon(data = world_map, aes(x = long, y = lat, group = group),
               fill = "grey95", colour = "grey80", linewidth = 0.1) +
  geom_point(data = Gb_wide,
             aes(x = Longitude, y = Latitude, colour = factor(Value.GB137,exclude = NULL)),
             shape = 16, size = 0.8, alpha = 0.7) +
  scale_shape_identity() +
  scale_colour_manual(values = c("#CC79A7", "#009E73","#F0E442", "black"),
                      na.value = "black",
                      labels = c("No", "Yes", "Not Known", "NAs")) +
  coord_fixed(1.3) +
  theme_void() +
  theme(legend.position = "none") +
  labs(title = "Grambank GB137: Clause-Final Negation")
# Creating Grambank GB137 map


comp_map <- WALS_map / Gb_map 
comp_map
# Displaying maps with patchwork
```
