---
title: "Global Negation Strategies: A Comparative Analysis"
description: |
  An investigation into the distribution of negation strategies using Grambank and WALS datasets.
preview: showcase1_files/figure-html/unnamed-chunk-1-1.png
author:
  - name: Holly Lloyd
date: 2026-03-27
output:
  distill::distill_article:
    toc: true
    toc_depth: 2
    self_contained: FALSE
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)

#install.packages("scales")
library(knitr)
library(scales)
library(patchwork)
library(sf)
library(tidyverse)
library(ggplot2)
library(broom)
# Loading in packages

Gb_values <- read.csv("values.csv")
Gb_languages <- read.csv("languages.csv")
Gb_parameters <- read.csv("parameters.csv")
wals_values <- read.csv("WALSvalues.csv")
wals_languages <- read.csv("WALSlanguages.csv")
# Loading data 
```
# Section 1: Grambank Areal and Genealogical Distribution

The following analysis examines whether negation strategies are primarily driven by geographic proximity (areal diffusion) or family lineage (genealogical inheritance). I focus on two key features GB107 (affixal/clitic negation) and GB137 (clause-final negation).

# 1.1 Chi-Squared Tests for Areal Effect

In order to evaluate the areal diffusion, I utilised Pearson's Chi-squared tests. These tests enable me to determine whether the presence of a negation strategy is significantly associated with specific geographic macroareas. 

```{r}
negation_params <- Gb_parameters[Gb_parameters $ ID %in% c("GB107", "GB137"),]
Gb_negation_values <- Gb_values[Gb_values $ Parameter_ID %in% c("GB107", "GB137"),]
# Filtering for specific parameters

Gb_negations <- merge(Gb_negation_values, Gb_languages,
                      by.x = "Language_ID", by.y = "ID")
# Integrating values with metadata

Gb_wide <- reshape(Gb_negations[, c("Language_ID", "Parameter_ID", "Value",
                                    "Family_name", "Macroarea", "Longitude", "Latitude")],
                   idvar = c("Language_ID", "Longitude", "Latitude"),
                   timevar = "Parameter_ID",
                   direction = "wide")
# Reshaping to wide and specifying column names

areal_table <- table(Gb_wide $ Value.GB137, Gb_wide  $ Macroarea.GB137)
areal_table_chi <- chisq.test(areal_table)
tidy(areal_table_chi) %>%
  mutate(p.value = scales::scientific(p.value)) %>%
  knitr::kable(caption = "Areal Association: GB137",
               col.names = c("Statistic", "P-Value", "DF", "Method"))

areal_table2 <- table(Gb_wide $ Value.GB107, Gb_wide  $ Macroarea.GB107)
areal_table_chi2 <- chisq.test(areal_table2)
tidy(areal_table_chi2) %>%
  mutate(p.value = scales::scientific(p.value)) %>%
  knitr::kable(caption = "Areal Association: GB107",
               col.names = c("Statistic", "P-Value", "DF", "Method"))
# Chi-squared to evaluation associations
```
# 1.2 Genealogical Stability

While areal factors are significant, many language families exhibit vertical stability. I defined this demonstration of strong inheritance as a language family with 100% internal consistency for the analysed features. 

```{r}
family_table <- table(Gb_wide $ Value.GB137, Gb_wide $ Family_name.GB137)
family_props <- prop.table(family_table, margin = 2)

strong_inheritance <- family_props["1", family_props["1", ] == 1.0]
as.data.frame(head(strong_inheritance)) %>%
  knitr::kable(caption = "Strong Inheritance Families: GB137",
               col.names = c("Family Name", "Proportion"))

family_table2 <- table(Gb_wide $ Value.GB107, Gb_wide $ Family_name.GB107)
family_props2 <- prop.table(family_table2, margin = 2)

strong_inheritance2 <- family_props2["1", family_props2["1", ] == 1.0]
as.data.frame(head(strong_inheritance2)) %>%
  knitr::kable(caption = "Strong Inheritance Families: GB107",
               col.names = c("Family Name", "Proportion"))
# Investigating genealogical stability
```

# 1.3 Feature Co-occurrence and Correlation

While the previous section focus on the external drivers of negation of geography and ancestry, it is also critical to examine the internal relationship in between these markers. This section investigates whether the use of affixal negation (GB107) and clause-final negation (GB137) are independent structural choices or if they exhibit a significant typological correlation.

The following Chi-squared test evaluates this dependency. A significant result would suggest that the presence of one negation strategy may historically or structurally constrain the development of the other, thus suggesting a deeper cross-linguistic pattern in how negation is encoded.

```{r}
feature_comp <- table(Gb_wide $ Value.GB107, Gb_wide $ Value.GB137)
feature_test <- chisq.test(feature_comp, simulate.p.value = TRUE)
tidy(feature_test) %>%
  knitr::kable(caption = "Correlation Test: GB137 vs. GB107",
               col.names = c("Statistic", "P-value", "Parameter", "Method"))
# Testing for correlation between features
```

# 1.4 Visualising Regional Variation

While Chi-squared tests provide statistical evidence for areal and genealogical trends, visualising these distributions may reveal a potential geographical 'shape' of negation. In this section I contrast clause-final negation (GB137) with affixal negation (GB107) and through plotting these as proportional bar charts across macroareas, it may be possible to determine if certain regions favour specific structural blueprints for negation.

```{r}
plotGB137 <- na.omit(Gb_wide[, c("Macroarea.GB137", "Value.GB137")])
plotGB137 <- plotGB137[plotGB137 $ Value.GB137 %in% c("0", "1"),]
#Preparing data for GB137 plot

p1 <- ggplot(plotGB137, aes(x = Macroarea.GB137, fill = as.factor(Value.GB137))) +
  geom_bar(position = "fill", width = 0.7) +
  scale_y_continuous(labels = percent) +
  scale_fill_manual(values = c("#56B4E9", "#0072B2"),
                    labels = c("Absent", "Present")) +
  labs(title = "Clause-Final Negation (GB137)",
       subtitle = "Syntactic Position:
       High regional variation suggests areal diffusion",
       y = "Percent of Languages", x = NULL, fill = "Presence") +
  theme_minimal(base_size = 11) +
  theme(legend.title = element_blank()) +
  coord_flip()
# Creating GB137 plot

plotGB107 <- na.omit(Gb_wide[, c("Macroarea.GB107", "Value.GB107")])
plotGB107 <- plotGB107[plotGB107 $ Value.GB107 %in% c("0", "1"),]
# Preparing data for GB107 plot

p2 <- ggplot(plotGB107, aes(x = Macroarea.GB107, fill = as.factor(Value.GB107))) +
  geom_bar(position = "fill", width = 0.7) +
  scale_y_continuous(labels = percent) +
  scale_fill_manual(values = c("#E69F00", "#D55E00"),
                    labels = c("Absent", "Present")) +
  labs(title = "Affixal/Clitic Negation (GB107)",
       subtitle = "Morphological Strategy:
       Broader distribution suggests genealogical stability",
       y = "Percent of Languages", x = NULL, fill = "Presence") +
  theme_minimal(base_size = 11) +
  theme(legend.title = element_blank()) +
  coord_flip()
# Creating GB107 plot

combined_plot <- p1 / p2 +
  plot_annotation(title = "Comparative Analysis of Negation Strategies",
                  caption = "Data: Grambank 2026") +
  plot_layout(guides = "collect")
combined_plot
# Combining plots using patchwork
```

The resulting visualisations highlight a clear divergence in how these two strategies are distributed globally. Clause-final negation exhibits significant peaks in specific macroareas like South America and Papuanesia, perhaps suggesting that this syntactic features may spread through language contact. However, affixal negation shows a more balanced, presence across most macroareas. This suggests that morphological negation is often a deeply embedded trait that is less prone to being lost or gained through geographical proximity and therefore more likely to be inherited vertically through language families.


# Section 2: WALS Structural Correlation

In contrast to the binary granularity of Grambank, the WALS dataset allows for the investigation into the dependency between clausal symmetry (113A) and syntactic Position (143G).

```{r}
wals_full <- merge(wals_values, wals_languages,
                   by.x = "Language_ID", by.y = "ID")
# Integrating WALS values with metadata

wals_113a <- wals_full[wals_full $ Parameter_ID == "113A", ]
wals_143g <- wals_full[wals_full $ Parameter_ID == "143G", ]
# Filtering for relevant parameters

wals_areal_table <- table(wals_113a $ Value, wals_113a $ Macroarea)
wals_chi <- chisq.test(wals_areal_table, simulate.p.value = TRUE)
tidy(wals_chi) %>%
  knitr::kable(caption = "Areal Association: 113A vs. Macroarea", 
col.names = c("Statistic", "P-Value", "DF", "Method"))

wals_areal_table2 <- table(wals_143g $ Value, wals_143g $ Macroarea)
wals_chi2 <- chisq.test(wals_areal_table2, simulate.p.value = TRUE)
tidy(wals_chi2) %>%
  knitr::kable(caption = "Areal Association: 1143G vs. Macroarea", 
col.names = c("Statistic", "P-Value", "DF", "Method"))
# Chi-squared for WALS features

wals_subset <- wals_full[wals_full $ Parameter_ID %in% c("113A", "143G"), ]
wals_wide <- reshape(wals_subset[, c("Language_ID", "Parameter_ID", "Value",
                                     "Family", "Macroarea")],
                     idvar = "Language_ID",
                     timevar = "Parameter_ID",
                     direction = "wide")
# Reshaping subset for correlation testing

wals_wide_clean <- na.omit(wals_wide[, c("Value.113A", "Value.143G")])
wals_wide_clean <- droplevels(wals_wide_clean)
# Cleaning data 
```
# 2.1 Strategy vs. Word Order

The contingency table below maps the frequency of specific strategy-order pairings. A significant chi-squared result suggests that certain syntactic positions favor more complex grammatical pathways.

```{r}
combined_wals_table <- table(wals_wide_clean $Value.113A, wals_wide_clean $ Value.143G)
# Creating contingency table

wals_corr <- chisq.test(combined_wals_table, simulate.p.value = TRUE)
# Performing Chi-squared

knitr::kable(combined_wals_table,
             caption = "Contingency Table: WALS 113A vs WALS 143G",
             col.names = c("V-Neg", "Neg-V"))
# Displaying table

tidy(wals_corr) %>%
  knitr::kable(
    caption = "Chi-squared Test Results: Correlation between Strategy and Order",
    col.names = c("Statistic", "P-Value", "DF", "Method"))
# Displaying Chi-squared
```

The above Chi-squared results for the WALS data 113A and 143G reveals no statistically significant correlations. Unlike the regional clustering observed in the Grambank data, these results suggest that the choice between symmetric and asymmetric negation strategies is not strictly dictated by basic word orders or the position of the negative element.

The independence may indicate that these structural dimensions evolve through different typological pressures. As while a language may shift its negative position due to contact or changes in word-order, the internal symmetry of its negation system appears  to remain as a distinct and independent structural trait.

# 2.2 Visualising Proportional Trends

While the previous Chi-squared test confirms a lack of statistical correlation, the following proportional plot allows for the inspection of how clausal symmetry (113A) is distributed across different syntactic positions (143G). Through this standardisation of frequencies to 100%, the relative prevalence of symmetric versus asymmetric negation strategies for both V-Neg and Neg-V word orders become observable.

The resulting  distribution illustrates the structural diversity of negation, as it demonstrates that neither syntactic position (113A or 143G) constrains a language into a specific morphological pathway. Noticeably, however, the data does reveal that symmetric strategies are only present within the Neg-V category, therefore potentially suggesting that these features are generally governed by independent evolutionary typological pressures, despite certain morphological configurations that may be restricted to specific word-order environments.

```{r, preview=TRUE}
wals_counts <- as.data.frame(table(wals_wide_clean $ Value.113A,
                                   wals_wide_clean $ Value.143G))
colnames(wals_counts) <- c("Strategy", "Order", "Frequency")
# Preparing data for bar chart


wals_counts $ Strategy <- factor(wals_counts $ Strategy,
                                 labels = c("Symmetric", "Asymmetric", "Asymmetric (Auxiliary)"))
wals_counts $ Order <- factor(wals_counts $ Order,
                              labels = c("V-Neg", "Neg-V"))
# Relabeling 

ggplot(wals_counts, aes(x = Order, y = Frequency, fill = Strategy)) +
  geom_bar(stat = "identity", position = "fill") +
  scale_y_continuous(labels = percent) +
  scale_fill_manual(values = c("#F0E442", "#E69F00", "#D55E00")) +
  labs(title = "Distribution of Negation Strategies by Word Order",
       subtitle = "Proportional analysis of WALS features 113A and 143G",
       caption = "Data: WALS 2026", 
       x = "Word Order (143G)", y = "Percentage of Languages", fill = "Strategy (113A)") +
  theme_minimal()
# Creating bar graph
```

# Section 3: Comparing WALS and Grambank Datasets

To synthesize the previous findings, I plotted the global distributions of WALS 113A (Clausal symmetry), alongside Grambank GB137 (Clause-final negation). Mapping these features allows me to move beyond abstract percentages and observe the actual geographic densities of negation strategies. These comparison visualisations highlight how specific structural choices cluster in particular regions, thus providing a final layer of evidence for the areal and genealogical trends discussed throughout this analysis.

```{r}
world_map <- map_data("world")

WALS_map <- ggplot() +
  geom_polygon(data = world_map, aes(x = long, y = lat, group = group),
               fill = "grey95", colour = "grey80", linewidth = 0.1) +
  geom_point(data = wals_113a, aes(x = Longitude, y = Latitude,
                                   colour = as.factor(Value)),
             size = 0.8, alpha = 0.7) +
  scale_colour_manual(values = c("#D55E00", "#E69F00", "#0072B2", "#56B4E9")) +
  coord_fixed(1.3) +
  theme_void() +
  theme(legend.position = "none") +
  labs(title = "WALS 113A: Symmetric and Asymmetric Negation")

Gb_map <- ggplot() +
  geom_polygon(data = world_map, aes(x = long, y = lat, group = group),
               fill = "grey95", colour = "grey80", linewidth = 0.1) +
  geom_point(data = Gb_wide,
             aes(x = Longitude, y = Latitude, colour = factor(Value.GB137, exclude = NULL)),
             shape = 16, size = 0.8, alpha = 0.7) +
  scale_shape_identity() +
  scale_colour_manual(values = c("#CC79A7", "#009E73", "#F0E442", "black"),
                      na.value = "black",
                      labels = c("No", "Yes", "Not Known", "NAs")) +
  coord_fixed(1.3) +
  theme_void() +
  theme(legend.position = "none") +
  labs(title = "Grambank GB137: Clause-Final Negation")

comp_map <- WALS_map / Gb_map
comp_map
```

The comparison between the WALS and Grambank maps illustrates the complementary nature of each dataset. While the WALS map (above) captures a broad typological overview of clausal symmetry with a focus on categorical variety, the Grambank map (below) provides a more dense, binary snapshot of syntactic positioning. In combination, they reveal that certain regions are hot-spots for diverse negation strategies. These maps confirm that while negation is a universal linguistic requirement, its structural implementation is deeply tied to the geographic and historical context of the world's language families.


*NOTES:  
  - Data retrieved from WALS and Grambank March 20, 2026  
  - Colour-blind accessible colours used (Okabe-Ito)*